Red Wine Quality by Pablo Rivero
========================================================

# INTRODUCTION
 Here I explore and analyze a data set that contains contains 1,599 red wines
with 11 variables on the chemical properties of the wine: "fixed.acidity",
"volatile.acidity", "citric.acid","residual.sugar", "chlorides",
"free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH", "sulphates",
"alcohol", and "quality". At least 3 wine experts rated the quality of each
wine, providing a rating between 0 (very bad) and 10 (very excellent). The goal
is to extract some conclusions about the relation between chemical properties
and the quality of the wines.

> I load some libraries

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```

```{r echo=FALSE}
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(car)
library(reshape)
library(plyr)
library(gridExtra)
#library(dplyr)
#library(tidyr)
```

>I load the DataFile

```{r echo=FALSE}
reds <- read.csv("wineQualityReds.csv")
```


# Univariate Plots Section #

The next is going to be a preliminary exploration of the dataset.
I have to understand the structure of the individual variables in the dataset. 
I first check the dimension, what variables I have, Min, Max, mean,medians
in order to have an initial sense

```{r echo=FALSE}
dim(reds)  #Dimension of our DataSet
str(reds)  #Gives me information about the type of the variable
summary(reds)  #Information about MIN,MAX, MEDIAN,MAX to understand the limits
theme_set(theme_minimal(12))
```

>Initial Observations

1- PH is between 2.74 and 4.01. Median is 3.31

2- Max quality value is 8. Min is 3. Median is 6

3- Alcohol is between 8.4% and 14.9%. 

4- Residual sugar range is large, between 0.9 and 15.5. But median is 2.2g/dm^3

Let's make some plots:


```{r echo=FALSE,fig.width=7, fig.height=12}
plot.hist <- function(x.var, df, bw){
  ggplot(df, aes_string(x = x.var)) +
    geom_histogram(binwidth = bw)
}

P1<-plot.hist('volatile.acidity', reds, 0.02) + xlab('volatile.acidity')
P2<-plot.hist('alcohol', reds, 0.1) + xlab('alcohol')
P3<-plot.hist('fixed.acidity', reds, 0.1) + xlab('fixed.acidity')
P4<-plot.hist('citric.acid', reds, 0.02) + xlab('citric.acid')
P5<-plot.hist('residual.sugar', reds, 0.1) + xlab('residual.sugar')
P6<-plot.hist('chlorides', reds, 0.005) + xlab('chlorides')
P7<-plot.hist('free.sulfur.dioxide', reds, 2) + xlab('free.sufur.dioxide')
P8<-plot.hist('density', reds, 0.0005) + xlab('density')
P9<-plot.hist('pH', reds, 0.02) + xlab('pH')
P10<-plot.hist('sulphates', reds, 0.02) + xlab('sulphates')
P11<-plot.hist('total.sulfur.dioxide', reds, 2) + xlab('total.sufur.dioxide')
P12<-plot.hist('quality', reds, 0.5) + xlab('quality')
grid.arrange(P1,P2,P3,P4,P5,P6,P7,P8,P9,P10,P11,P12,ncol=2)
```


- Residual sugar, chlorides and sulfates have similar distributions

- pH and density have a very similar distribution. Correlation can be important
here.

One of the main interests in this study is to understand the chemical properties
that produce a great quality wine. 

I decided to break quality in 3 ranges and consider 7 and 8 high quality (since
only 18 wines have quality) (0-5], (5-6], and (7-10] and check some chemical
properties like pH and alcohol.


```{r echo=FALSE}

reds$quality.bucket<-cut(reds$quality,c(0,5,6,10))

plot.hist2 <- function(x.var, df, bw){
  ggplot(df, aes_string(x = x.var)) +  facet_wrap(~quality.bucket)+
    geom_histogram(binwidth = bw)
}

P1<-plot.hist2('pH', reds, 0.02) + xlab('pH')
P2<-plot.hist2('sulphates', reds, 0.02) + xlab('sulphates')
P3<-plot.hist2('density', reds, 0.0005) + xlab('density')
P4<-plot.hist2('alcohol', reds, 0.1) + xlab('alcohol')

grid.arrange(P1,P2,P3,P4,ncol=2)
summary(reds$quality.bucket)
```

Only 13.6% of wines have 6 or more quality value 
All quality wines have similar amount of pH. Not important factor. Same for
sulphates and density. However, No good wines (6,10 range) have less than 10%
of alcohol. 

In thee following plots I want to understand the desviation of the
data for some variables using the boxplot function

```{r echo=FALSE}
grid.arrange(ggplot(reds, aes( x = 1, y = alcohol ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(reds, aes( x   = alcohol  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
```    

There are some points that could be considered outliers at >=14% alcohol or
below 9% of alcohol. We should discuss if consider these datapoints when doing
stadistics.
Most of the wines are between 9 and 11

```{r echo=FALSE}
grid.arrange(ggplot(reds, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(reds, aes( x   = residual.sugar  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
```                   


Most of the wines have between ~1.7 and 3.4 of residual sugar. In this case it
looks like there are a huge amount of outliers. Probablyy we should not 
consider all wines above 8. Large deviations

```{r echo=FALSE}
grid.arrange(ggplot(reds, aes( x = 1, y = pH ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(reds, aes( x   = pH  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
``` 
  
          
Most of the wines have between 3 and 3.6 of pH. In this case the deviation is
not as high as the residual sugar. Some outliers are below 2 or above 4.
                   
### What is the structure of your dataset?

Dataset is made by 1599 observables and 13 variables

### What is/are the main feature(s) of interest in your dataset?

The quality of the wine

### What other features in the dataset do you think will help support your \

The relation between the properties of the wine with the quality value.

### Did you create any new variables from existing variables in the dataset?

Yes: quality.bucket in order to group better the quality of the wines.

### Of the features you investigated, were there any unusual distributions?

Unusual distributions were: citric.acid and free.SO2
Normal distributions: density, pH, and vlatile.acidity
Negative: sugar, total SO2, alcohol and sulphates

### Did you perform any operations on the data to tidy, adjust, or change the 
###form of the data? If so, why did you do this?

Yes. I created reds_short subset data to explore different bivariate plots and 
correlations with less amount of variables (next section)



# Bivariate Plots Section

Let's check with ggpairs different Bivariate plots to have an initial sense of
the highest correlations between variables in this DataSet

```{r echo=FALSE,fig.width=10, fig.height=7}
set.seed(2183)
reds_short<-subset(reds,select=c(alcohol,density,sulphates,residual.sugar,
                    pH,total.sulfur.dioxide,free.sulfur.dioxide,
                    citric.acid,fixed.acidity,volatile.acidity,
                    quality))
ggpairs(reds_short, 
        lower = list(continuous = wrap("points", alpha = 0.3,    size=0.1)),
        upper = list(combo = wrap("dot", alpha = 0.4, size=0.2,
                                  params=list(corSize=12))),
        title = "Wines")
```


Notes about correlations:
1: There is a good correlation (0.668) between fixed acidity and density. 

2: There is a good negative correlation (-0.683) between fixed acidity and pH.

3: There is a good correlation (0.668) between Total SO2 and free SO2. 

4: There is a good correlation (0.672) between fixed acidity and citric.acid.

Let's see some of this relations with their regresion lines



```{r echo=FALSE}
ggplot(aes(x=fixed.acidity,y=density),data=reds)+
  geom_jitter(alpha=1/10)+geom_smooth(method = 'lm',color='red')
cor.test(reds$fixed.acidity,reds$density,method='pearson')
```


Density vs fixed acidity shows a R=0.668. There is still some big dispersion
with datapoints that should be removed (density=0.99,acidity=8). Bur due to the
large amount of data those point would not change to much the value of R.


```{r echo=FALSE}
ggplot(aes(x=fixed.acidity,y=pH),data=reds)+
  geom_jitter(alpha=1/10)+geom_smooth(method = 'lm',color='red')
cor.test(reds$fixed.acidity,reds$pH,method='pearson')
```


There is a good negative correlation (-0.683) between fixed acidity and pH.

```{r echo=FALSE}
ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide),data=reds)+
  geom_jitter(alpha=1/10)+geom_smooth(method = 'lm',color='red')
cor.test(reds$free.sulfur.dioxide,reds$total.sulfur.dioxide,method='pearson')
```


There is a good correlation (0.668) between Total SO2 and free SO2. 

Let's focus now in the chemical properties of a good wine in comparison with 
other quality wine. 


```{r echo=FALSE}
ggplot(aes(factor(quality.bucket), 
            alcohol), 
        data = reds) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```


Alcohol increases with the quality of the wine

```{r echo=FALSE}
ggplot(aes(factor(quality.bucket), pH), data = reds) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```


pH remains between 3.3 and 3.8 for all quality wines.

```{r echo=FALSE}
ggplot(aes(factor(quality.bucket), chlorides), data = reds) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```


Less dispersion can be seen on chlorides as we go for high quality wines. 
Keeping them close to 0.1


It was interesting that just a few wines had below 10% of alcohol so I wanted
to know more:

```{r echo=FALSE}
library(dplyr)
reds$alcohol.bucket<-cut(reds$alcohol,c(8,10,12,14,15))
ggplot(data=reds, aes(x=pH,fill=alcohol.bucket)) +
  scale_fill_brewer(palette = "Reds")+
  facet_wrap(~quality.bucket)+
  geom_histogram()
by(reds$alcohol,reds$quality,summary)
```


While there is not too much wines with more than 12% of alcohol in the low
range of quality, there are not below 10% in the high quality level. 

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I used a multiscatter plot to look for relationships between all chemicals.
I also looked individually how each chemical is related with the quality of the
wine. I found that we can only get some hints about how these chemicals can
produce a good quality wine and it is more the sum of these chemicals. However,
I found that the degree of alcohol increases with the quality of wine.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

1: There is a good correlation (0.668) between fixed acidity and density. 

2: There is a good negative correlation (-0.683) between fixed acidity and pH.

3: There is a good correlation (0.668) between Total SO2 and free SO2. 

4: There is a good correlation (0.672) between fixed acidity and citric.acid.

5: Quality and level of alcohol

### What was the strongest relationship you found?

The fixed acidity with the pH

# Multivariate Analysis

```{r echo=FALSE}
ggplot(aes(x=fixed.acidity,y=density,color=pH),data=reds)+
  scale_color_gradientn(colours=blues9)+
  facet_wrap(~quality.bucket)+geom_jitter()+
  geom_smooth(method = 'lm',color='red')

```


Good correlation between pH and fixed acidity. Where low pH correspond with 
low fixed acidity, and when density increases, fixed acidity too, but not pH.

```{r echo=FALSE}
mid<-mean(reds$pH)
ggplot(aes(x=alcohol,y=density,color=pH),data=reds)+
    facet_wrap(~quality.bucket)+geom_jitter()+
  geom_smooth(method = 'lm',color='red')+ylab('Density ((g/cm^3))')+
  ylab('Alcohol (%)')
```


Good correlation between alcohol and density. This correlation looks better
for more quality wines although still a lot of dispersion.
High alcoholic wines usually have more pH. 


By considering a wine with the following qualities:
alcohol>10%
sulphates>0.70
pH<3.3
chlorides<0.075
volatile.acidity<0.4
total.sulfur.dioxide<40

```{r echo=FALSE}
test=subset(reds,alcohol>9.5 & sulphates>0.70 & pH<3.3 & chlorides<0.075 &
              volatile.acidity<0.4 & total.sulfur.dioxide <40)
summary(reds$quality.bucket)
summary(test$quality.bucket)

```


We go from 13.6% of 7 or 8 quality wines to 67.5%

Based on this I create a model to predict the quality of a wine based on
the amount of the different chemicals it has.

```{r echo=FALSE}
m1<-lm(quality ~ alcohol,data=reds)
m2<-update(m1,~.+ pH)
m3<-update(m2,~.+ sulphates)
m4<-update(m3,~.+ volatile.acidity)
m5<-update(m4,~.+ total.sulfur.dioxide)
mtable(m1,m2,m3,m4,m5)
```


### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

-Low fixed acidity have lower pH as we supposed

-A high quality wine should have: low pH, low amount of chlorides,
volatile.acidity, density and high amount of citric acidity, alcohol degree,
and sulphates.

### Were there any interesting or surprising interactions between features?

Alcohol is directly related with quality of wine.
Fixed acidity is related with pH

### OPTIONAL: Did you create any models with your dataset? Discuss the \

Create a model to predict the quality of a wine based on the amount
of the chemical variables but R^2=0.347 (too low). There is too much dispersion
to arrive to an a conclusion like this based on the chemicals and the different
taste of 3 persons.
------


# Final Plots and Summary

### Plot One
```{r echo=FALSE}
library(dplyr)
reds$alcohol.bucket<-cut(reds$alcohol,c(8,10,12,14,15))
ggplot(data=reds, aes(x=pH,fill=alcohol.bucket)) +
  scale_fill_brewer(palette = "Reds")+
  facet_wrap(~quality.bucket)+
  geom_histogram()+
  geom_vline(xintercept = 3.3, 
                color = "black", size=1)+
  ggtitle('Amount of wines as a function of pH and alcohol')+
  theme(plot.title = element_text(size = 20, face = "bold"))+
  ylab('Number of wines')
by(reds$alcohol,reds$quality.bucket,summary)
by(reds$pH,reds$quality.bucket,summary)
```



### Description One
The amount of alcohol degree is important when considering a high quality wine.
As we can see in the previous figure. Practicaly no wines with more than 12% of
alcohol have qualities over 5. Being here, the wines with less than 10% the
are the most. On the other side, wines with more than 12% of alcohol are 
abundant on quality wines of 7 or 8 values while there are almost no wines
with lss than 10%. 
pH=3.3 (black solid line) is kept more or less in the middle of all 
distributions. We can see some displacement towards a loer pH when we go to 
high quality wines.

### Plot Two

```{r echo=FALSE}
ggplot(aes(factor(quality), 
            alcohol), 
        data = reds) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  ggtitle('Alcohol vs quality')+ylab('Alcohol(%)')+xlab('Quality')+
  theme(plot.title = element_text(size = 20, face = "bold"))
```


### Description Two

In this plot we can see how evolves the median for alcohol as a function
of quality. A high quality wine should have high level of alcohol in general.
However, dispersion is big.

We go from 13.6% of 7 or 8 quality wines to 67.5%

### Plot Three

```{r echo=FALSE}
ggplot(aes(x=fixed.acidity,y=density,color=pH),data=reds)+
  scale_color_gradientn(colours=blues9)+
  facet_wrap(~quality.bucket)+geom_jitter()+
  geom_smooth(method = 'lm',color='red')+
  ggtitle('Fixed acidity vs density and pH per quality wine')+
  theme(plot.title = element_text(size = 18, face = "bold"))+
  ylab('Density(g/cm^3)')+
  xlab('Fixed Acidity (g/dm^3)')
```


### Description Three

Low fixed acidity have high pH (blue dots) as we could have thought. 
Besides, from the previous figure we can also see that density increases
with the fixed acidity increment. The correlation between these two
variables is the highest I found in the DataSet.

# Reflection

I found several correlations between different chemicals on wine. I could
understand, in general, what of these chemicals can have direct relationships
with the quality of the wine. My struggles here were that, since I do not know
too much of these chemicals (more literature research could have worked) I 
started the analysis quite blind. Only alcohol, pH and quality had sense for me
Also, I wanted to find a model to predict the quality of the wine based on the
amount of chemicals but it turns out that the correlation of that function was
quite bad (R~0.6)
-I think more persons to qualify the wine would have given better results.
-Other variables like age of the wine could also be important.


